Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-03-27 Publications/2015-NeuroImage/03-introduction.tex
reworking intro

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}

\Gls*{dmri} data of the brain present severe distortions, directly related to the
  effort in minimizing acquisition time.
In order to quickly probe the diffusion process in sufficiently different orientations,
  \gls*{dmri} are acquired with \gls*{epi} sequences, which minimize the bandwidth of the \gls*{pe}
  direction.
This limitation originates a warping of the reconstructed image in regions affected by
  small deviations from the main magnetic field $B_0$.
One important source of these deviations is the discontinuity of magnetic susceptibility across
  the imaged object.
In fact, the distortion is the effect of the underlaying misregistration between correct and
  warped data.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe and the temporal
  bone of both hemispheres.
Several approaches have been proposed to estimate the distortion and correct
  for the artifact.
The first solution consisted on estimating the deviations from $B_0$ with field mapping
  techniques \citep{andersson_modeling_2001}.
A second family of methods require an extra \gls*{epi} acquisition, but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient \gls*{pe}
  increments \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed nonlinear registration of the \emph{b0}
  image of the \gls*{dmri} to an undistorted \gls*{t2} image of
   the same subject.

The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.
Currently, a broad range of applications make use of whole-brain tractography, such as
  connectome extraction \citep{craddock_imaging_2013} or tract-based spatial statistics
  \citep{smith_tractbased_2006}.
These applications generally require the segmentation of the \gls*{dmri} data, also
  hindered by distortions.
One way to identify the regions of interest is extracting them from an appropriate 
  \gls*{t1} of the same subject and map them into diffusion space,
  in a sort of atlas-based segmentation.

\paragraph{State of the art}\label{sec:state_of_art}
  Active contours have been successfully integrated in image registration methods
  for applications such as longitudinal object tracking \citep{paragios_level_2003} and
  atlas-based segmentation \citep{gorthi_active_2011}.
\cite{bertalmio_morphing_2000} proposed morphing active-contours for object tracking in 
  series of 2D images.
Shortly, \cite{yezzi_variational_2003} presented the first method including a full solution to
  the registration problem with an affine transformation supporting the mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration framework using level sets and only
  one \gls*{pde} for first time.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2003}
  to nonlinear registration implementing a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level-set function for the registration of several active contours, in the application
  of atlas-based segmentation.

\cite{greve_accurate_2009} presented a relevant registration method called \emph{bbregister},
  now included in \emph{FreeSurfer} \citep{fischl_freesurfer_2012}.
The method is a standard choice to map surfaces extracted in the \gls*{t1} image to the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value.} of the \gls*{dmri} dataset with two limitations.
First, the transform is affine and therefore it does not cope with the nonlinear distortions.
Second, \emph{bbregister} looks for intensity steps in the image, using active contours with edges.
In practice, the target \emph{b0} does not present clear frontiers of intensity between
  interfacing tissues, what limits \emph{bbregister} to mapping the cortical surface only.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely and coregistration of anatomical
  images are difficult for several reasons.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  interfacing voxels due to the low resolution of images (generally around
  \isores{2.2}mm).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  use for direct segmentation.
The contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Finally, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of presenting comprehensively reported geometrical distortion and 
  signal losses \citep{jezzard_correction_1995} that hinder the segmentation by
  registering an anatomical image as source of same-subject prior information.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map at values around 0.7.
The approach is unreliable for the high variability of \gls*{fa} and its
  decay with the increase of incoherence between fibers.
\cite{zhukov_level_2003} proposed active contours with edges represented
  by level sets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based level-sets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targets the definition of appropriate features to segmentation,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  graph-cuts \citep{han_experimental_2009}, and volume fraction modeling
  \citep{kumazawa_improvement_2013}.

\cite{saad_new_2009} used Pearson correlation to perform linear registration
  of the \gls*{t1} image to the \emph{b0} of the \gls*{dmri}.
Another linear registration method is \emph{bbregister} \citep{greve_accurate_2009},
  that uses active contours extracted from the \gls*{t1}\footnote{using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}} to look for intensity boundaries in the \emph{b0}
  image.
In practice, the target \emph{b0} does not present clear frontiers of intensity between
  interfacing tissues, what limits \emph{bbregister} to mapping the cortical surface only.
Since the distortions found in \gls*{dmri} are nonlinear, \emph{bbregister} excludes
  from the boundary search those regions typically warped by the artifact.
This tool has become the standard method for its proven robustness.
Nonlinear registration has been successfully performed between \gls*{t2} and \emph{b0}
  images for their similarity, but uniquely as a way to correct for distortions
  \citep{kybic_unwarping_2000,studholme_accurate_2000,wu_comparison_2008,tao_variational_2009}.
Further registration of \gls*{t1} and \gls*{t2} images was still required to map the anatomical
  information (and the \gls*{wm} segmentation) into \gls*{dmri} space.

To jointly solve registration and segmentation problems such as longitudinal object
  tracking \citep{paragios_level_2003} and atlas-based segmentation 
  \citep{gorthi_active_2011}, active contours have been successfully integrated 
  in image registration methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended an existing linear registration \cite{yezzi_variational_2003}
  to the nonlinear case implementing a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level-set function for the registration of several active contours, in the application
  of atlas-based segmentation.
